mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-21 Draft3.Rmd
operators and functions

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + {n*o}^{p + q})
mathout(term)
```

todo: parentheses around -infty, see also $(no)^{p+q}$ below.

@@

Note that for the standard operators with known precedence, \pkg{mathml} is able
to detect if parentheses are needed; for example, a parenthesis is automatically
put around `g + h` in the `nodot`-example and around `n*o` above. However, we
also note an unecessary parenthesis around `l + m` above. This parenthesis
results is a consequence of `quote(a^(b + c))` actually producing a nested
R\ call of the form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

This feature is unfortunate because the extra parenthesis around `b + c` is not
needed and yields to the extra parenthesis. The preferred result is obtained by
using the functional representation `quote('^'(k, l + m))` of the power, or, as
a workaround, curly braces (see last summand above).

@@

As can be seen, \pkg{mathml} reminds the user that it is not possible to change
the precedence of custom operators in R\, and that expressions with such
operators are evaluated strictly from left to right. As in the
previous example, the solution is to either use the functional form of the
operator or a curly brace to enforce the correct operator precedence.

@@

More operators are shown in Table \@ref(tab:custom-operators).

```{r custom-operators, echo=FALSE}
op <- list(
  cross=quote(x %*% y),
  dot=quote(x %.% y),
  kronecker=quote(X %x% Y),
  'integer division'=quote(x %/% y),
  and=quote(A & B),
  vel=quote(A | B),
  xor=quote(xor(A, B)),
  not=quote(!A),
  equal=quote(X == Y),
  assign=quote(X <- Y),
  different=quote(x != y),
  tilde=quote(X ~ Y),
  approx=quote(x %~~% y),
  equiv=quote(x %==% y),
  congr=quote(x %=~% y),
  prop=quote(x %prop% y),
  'in'=quote(a %in% A),
  intersect=quote(intersect(A, B)),
  union=quote(union(A, B)),
  intersect=quote(is.null(intersect(A, B)))) # todo: check if duplicate

m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))

t <- cbind(Operator=op, Output=m)
knitr::kable(t, caption="Custom operators in mathml", row.names=FALSE, escape=FALSE)
```

@@

The second example may be particularly of use when we deal with multiple conditions (see \@ref(case-study)). Here, for instance, we present a trivial one with different relation signs.

```{r, results='asis'}
f <- function(x, a)
{
    if(a == 0L) return( 0L );
    if(a <= 1L) return(-Inf);
    if(a > 1L) return(+Inf)
}

Note that you may also assign a name to the function. The default name is in fact the greek letter lambda, but with `attributes()$name` the name costumization is performed smoothly.

```{r, results='asis'}
attributes(f)$name = quote(f)
mathout(f)
```

A few more elements are displayed in
Tables \@ref(tab:basic-elements) and \@ref(tab:arrows).

```{r basic-elements, echo=FALSE}
op1 <- list(
  lchoose=quote(lchoose(n, k)),
  sqrt=quote(sqrt(x)),
  mean=quote(mean(xy)),
  absolute=quote(abs(x))) # todo: check if duplicate
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

@@

>>>>

```{r, results="asis"}
term <- quote(a - ((b + c)) - d*e + f*(g + h) + i/j + k^(l + m) + (n*o)^{p + q})
mathout(term)
```

@@

Note that for the standard operators with known precedence, \pkg{mathml} is able
to detect if parentheses are needed; for example, a parenthesis is automatically
put around `g + h` in the `nodot`-example. However, we also note an unecessary
parenthesis around `l + m` above. This parenthesis is a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

This feature is unfortunate because the extra parenthesis around `b + c` is not
needed and yields to the extra parenthesis. The preferred result is obtained by
using the functional representation `quote('^'(k + l, m + n))` of the power, or,
as a workaround, curly braces (see last summand above).

@@

This is a reminder that it is not possible to change the precedence of custom
operators in R\, and that expressions with such operators are evaluated strictly
from left to right. As in the previous example, the solution is to either use
the functional form of the operator or a curly brace to enforce the correct
operator precedence.

@@

More operators are shown in Table \@ref(tab:custom-operators), following the
suggestions by Murrell and Ihaka [-@murrell2000].

```{r custom-operators, echo=FALSE}
op1 <- list(
  cross=quote(x %*% y),
  dot=quote(x %.% y),
  kronecker=quote(X %x% Y),
  'integer division'=quote(x %/% y),
  and=quote(A & B),
  vel=quote(A | B),
  xor=quote(xor(A, B)),
  not=quote(!A),
  equal=quote(X == Y),
  assign=quote(X <- Y))

m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

op2 <- list(
  different=quote(x != y),
  tilde=quote(X ~ Y),
  approx=quote(x %~~% y),
  equiv=quote(x %==% y),
  congr=quote(x %=~% y),
  prop=quote(x %prop% y),
  'in'=quote(a %in% A),
  intersect=quote(intersect(A, B)),
  union=quote(union(A, B)),
  intersect=quote(is.null(A)))

m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2)
knitr::kable(t, caption="Custom operators in mathml", 
             row.names=FALSE, escape=FALSE)
```

@@

All functions from R::base are supported, a subset is shown in 
Tables \@ref(tab:basic-elements).

```{r basic-elements, echo=FALSE}
op1 <- list(
  sin=quote(sin(x)),
  cosh=quote(cosh(x)),
  tanpi=quote(tanpi(alpha)),
  log=quote(log(p)),
  log1p=quote(log1p(x)),
  exp=quote(exp(x)),
  expm1=quote(expm1(x)),
  lchoose=quote(lchoose(n, k)),
  sqrt=quote(sqrt(x)),
  mean=quote(mean(xy)),
  absolute=quote(abs(x)))
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

@@

For self-written functions, matters are a bit more complicated because the
underlying R package `rolog` [@rolog] cannot easily access the function name,
hence a default is chosen.

```{r, results='asis'}
f <- function(x)
{
  if(x == 0) return(0L);
  if(x <= 0) return(-1L);
  if(x > 0) return(1L)
}

Custom names can be given via attributes.

```{r, results='asis'}
attributes(f)$name = quote(sgn)
mathout(f)
```
